---
title: "PEhub: detecting multi-way enhancer hubs"
author: "Yidan Sun"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PEhub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Introduction

Chromatin interaction assays such as HiChIP record enhancer–promoter contacts one
pair at a time. Within a single cell, proximity ligation joins a promoter to only
one of its enhancers, so a hub of several enhancers engaging the same promoter is
fragmented across the cell population into overlapping pairwise contacts.

`r Biocpkg("PEhub")` reconstructs those hubs from the pairwise data.

# Installation

```{r}
remotes::install_github("yourname/PEhub")
```

No Python configuration is required. The Leiden backend is provisioned
automatically by `r Biocpkg("basilisk")` on first use.

# Preparing the annotation

`pehub_prepare_interactions()` needs a `GRanges` of transcription start sites
with a `gene_id` metadata column.

```{r}
library(PEhub)

gtf <- rtracklayer::import("genes.gtf.gz")
tss <- GenomicRanges::resize(gtf[gtf$type == "transcript"],
                             width = 1, fix = "start")
```

Alternatively, if you have the relevant annotation packages installed:

```{r}
tss <- build_tss_from_txdb()   # defaults to hg38 knownGene
```

# Running the whole pipeline

```{r}
result <- pehub_run(
  loop_file     = "significant_interactions.bedpe",
  loop_file_all = "all_interactions.bedpe.gz",
  outdir        = "results/",
  tss_input     = tss,
  filename      = "GM12878",
  workers       = 10
)

head(result$high_confidence)
```

# Single-cell / pooled low-depth input

PEhub's background model, weighting schemes and null-model permutation testing
were developed and validated on deeply sequenced bulk HiChIP and Pore-C data
(paired human-heart H3K27ac HiChIP samples; GM12878 Pore-C), where "pooled"
effectively means millions of input cells sequenced to high depth in a single
library. Nothing about PEhub's reconstruction logic is specific to that
regime — it exists precisely to recombine the single enhancer-promoter
contact each individual cell shows (proximity ligation only ever joins a
promoter to one of its enhancers at a time) into a population-level hub — but
the method's statistical calibration assumes the pool is large and deep.

If your `loop_file_all` instead comes from pooling a modest number of single
cells or nuclei (for example a single-cell/single-nucleus Hi-C-type assay,
one donor or one condition at a time), run a pre-flight check before trusting
hub calls, especially at longer genomic distances:

```{r}
pehub_check_power("all_interactions.bedpe.gz", n_cells = 66)
```

This prints, and returns, a flag:

* `"ok"` — depth looks adequate across all tested distance bins.
* `"low_power"` — total depth is below a threshold at which hub sizes were
  observed to collapse to the `k_min` floor in real single-donor calibration
  data, rather than forming a real distribution. Treat hub sizes and
  significance calls with caution, and avoid comparing hub counts/sizes
  across samples that differ substantially in this total.
* `"long_range_unreliable"` — overall depth is adequate, but one or more
  distance bins beyond ~50kb (the range most relevant to real enhancer-hub
  architecture, as opposed to short-range/proximal contacts) have too few
  reads for the background model to be numerically reliable there, even
  though it may still be fine at short range.

`pehub_prepare_interactions()` (and therefore `pehub_run()`) calls this check
automatically by default (`check_power = TRUE`) and attaches the result as
`$power_check` on its return value — set `check_power = FALSE` to skip it,
e.g. for input you already know is deeply sequenced bulk data.

```{r}
prep <- pehub_prepare_interactions(
  loop_file     = "significant_interactions.bedpe",
  loop_file_all = "all_interactions.bedpe.gz",
  outdir        = "results/",
  tss_input     = tss,
  filename      = "donor1",
  n_cells       = 66
)
prep$power_check$overall_flag
```

This check does not change any downstream computation — it is a diagnostic,
not a correction. A `"low_power"` or `"long_range_unreliable"` flag does not
mean PEhub refuses to run; it means the resulting hub calls should be
interpreted with the same caution a shallow bulk sample would warrant, and
compared across samples only after confirming they are in a similar depth
regime (or after explicit depth-matching upstream).

# Running the stages separately

Each stage returns its key objects, so intermediate results can be inspected
without loading any files.

## Stage 1: prepare interactions

```{r}
prep <- pehub_prepare_interactions(
  loop_file       = "significant_interactions.bedpe",
  loop_file_all   = "all_interactions.bedpe.gz",
  outdir          = "results/",
  tss_input       = tss,
  filename        = "GM12878"
)
```

## Stage 2: detect candidate hubs

```{r}
hubs <- pehub_detect_hubs(
  outdir        = "results/",
  filename      = "GM12878",
  weight_method = "bin_log_ratio_sig",
  resolution    = 1.0,
  k_min         = 3
)

nrow(hubs$hub1)   # one candidate hub per promoter
```

## Stage 3: evaluate hubs

```{r}
ev <- pehub_evaluate_hubs(
  outdir      = "results/",
  filename    = "GM12878",
  B_pvalue    = 1000,
  B_stability = 10
)

summary(ev$hub1$hub_p_value_global)
summary(ev$hub1$reproducibility_rate)
```

## Stage 4: classify and export

```{r}
out <- pehub_export_hubs(
  outdir           = "results/",
  filename         = "GM12878",
  pvalue_cutoff    = 0.05,
  stability_cutoff = 0.5
)

out$files
```

# Choosing a weighting scheme

Eleven schemes are available, combining contact count, statistical significance
and genomic distance in different ways.

```{r}
pehub_weight_schemes()
```

The default, `bin_log_ratio_sig` ("Distance-adjusted log-enrichment"), scores each
contact by the log ratio of its observed count to the median count at the same
genomic distance, scaled by significance. This makes the weight dimensionless and
comparable across distance scales.

# Session info

```{r, eval=TRUE}
sessionInfo()
```
