---
title: "PEhub: detecting multi-way enhancer hubs"
author: "Yidan Sun"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PEhub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Introduction

Chromatin interaction assays such as HiChIP record enhancer–promoter contacts one
pair at a time. Within a single cell, proximity ligation joins a promoter to only
one of its enhancers, so a hub of several enhancers engaging the same promoter is
fragmented across the cell population into overlapping pairwise contacts.

`r Biocpkg("PEhub")` reconstructs those hubs from the pairwise data.

# Installation

```{r}
remotes::install_github("yourname/PEhub")
```

No Python configuration is required. The Leiden backend is provisioned
automatically by `r Biocpkg("basilisk")` on first use.

# Preparing the annotation

`pehub_prepare_interactions()` needs a `GRanges` of transcription start sites
with a `gene_id` metadata column.

```{r}
library(PEhub)

gtf <- rtracklayer::import("genes.gtf.gz")
tss <- GenomicRanges::resize(gtf[gtf$type == "transcript"],
                             width = 1, fix = "start")
```

Alternatively, if you have the relevant annotation packages installed:

```{r}
tss <- build_tss_from_txdb()   # defaults to hg38 knownGene
```

# Running the whole pipeline

```{r}
result <- pehub_run(
  loop_file     = "significant_interactions.bedpe",
  loop_file_all = "all_interactions.bedpe.gz",
  outdir        = "results/",
  tss_input     = tss,
  filename      = "GM12878",
  workers       = 10
)

head(result$high_confidence)
```

# Running the stages separately

Each stage returns its key objects, so intermediate results can be inspected
without loading any files.

## Stage 1: prepare interactions

```{r}
prep <- pehub_prepare_interactions(
  loop_file       = "significant_interactions.bedpe",
  loop_file_all   = "all_interactions.bedpe.gz",
  outdir          = "results/",
  tss_input       = tss,
  filename        = "GM12878"
)
```

## Stage 2: detect candidate hubs

```{r}
hubs <- pehub_detect_hubs(
  outdir        = "results/",
  filename      = "GM12878",
  weight_method = "bin_log_ratio_sig",
  resolution    = 1.0,
  k_min         = 3
)

nrow(hubs$hub1)   # one candidate hub per promoter
```

## Stage 3: evaluate hubs

```{r}
ev <- pehub_evaluate_hubs(
  outdir      = "results/",
  filename    = "GM12878",
  B_pvalue    = 1000,
  B_stability = 10
)

summary(ev$hub1$hub_p_value_global)
summary(ev$hub1$reproducibility_rate)
```

## Stage 4: classify and export

```{r}
out <- pehub_export_hubs(
  outdir           = "results/",
  filename         = "GM12878",
  pvalue_cutoff    = 0.05,
  stability_cutoff = 0.5
)

out$files
```

# Choosing a weighting scheme

Eleven schemes are available, combining contact count, statistical significance
and genomic distance in different ways.

```{r}
pehub_weight_schemes()
```

The default, `bin_log_ratio_sig` ("Distance-adjusted log-enrichment"), scores each
contact by the log ratio of its observed count to the median count at the same
genomic distance, scaled by significance. This makes the weight dimensionless and
comparable across distance scales.

# Session info

```{r, eval=TRUE}
sessionInfo()
```
